Unifying Gompertzian growth with the communicable disease spreading paradigm

### 2022-05-26 PNAS/PNAS-main.tex
move figures inside pnas folder
update letter w refereese + slight mod to manuscript

\affil[a]{Abel X\\Enerhauggata 1\\0651, Oslo}
\affil[b]{University of Oslo}

% Please give the surname of the lead author for the running footer
\leadauthor{Haugen} 
\graphicspath{{../figures/}}
% Please add here a significance statement to explain the relevance of your work
\significancestatement{
We seek to connect the communicable disease paradigm with the recently observed mortality patterns of the initial Coronavirus pandemic in March-April 2020 that have been shown by many to exhibit Gompertzian growth. In doing this, we show that this type of growth pattern is incompatible with traditional communicable disease spreading models, but rather with a model through which a ubiquitus and instanteneous stressor elicits a mutual stress response.
The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" suggesting a paradigm shift away from that of a purely communicable paradigm to a hybrid where the environment plays a bigger role.}

@@

\subsection{Unification through a modified SIR model}
However, it is possible to reconcile the SIR model with the Gompertz curve. 
Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
\begin{equation}
\label{eq:reverseNetworkSIR}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(1-x_j) \quad \forall i.
\end{equation} 
This crucial change is based on the environmental stressor hypothesis rather than a communicable disease assumption.
Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
\begin{equation}
\sum_i a_{ij} = 1 \quad \forall j.
\end{equation} 
If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
\begin{equation}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
\end{equation}

@@

\begin{equation}
\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
\end{equation}
As we are interested in the aggregate macroscopic behavior, we take averages in the log domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
\begin{equation}
\frac{d}{{dt}} \ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i{x_i}\right]^{\frac{1}{N}} - \alpha.
\end{equation}
If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
An interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above. One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain (\ref{eq:medianGomp}),
\begin{equation}
\frac{d}{dt}\ln{M(t)} = -\beta \ln{M(t)}.
\end{equation}

@@

\matmethods{
\subsection*{Estimating model parameters}
The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (See SI appendix for details). 
Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths. 
Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.

On statistical grounds, one could argue that the daily mortality counts follow a Poisson distribution. However, in a growth setting, one might expect heteroskedasticity inherent in the biological process where e.g. growth has less variance in the later stages, perhaps dependent on proximity to a stationary state. This latter hypothesis could be incorporated in the stochastic framework presented in the text, but to keep the scope of this paper confined we use a simple non-linear least squares minimization as our results are not meaningfully different compared with a more complex variance model.

>>>>

\affil[a]{Independent Researcher, Oslo}
\affil[b]{Independent Researcher, Oslo}

% Please give the surname of the lead author for the running footer
\leadauthor{Haugen}

% Please add here a significance statement to explain the relevance of your work
\significancestatement{
We seek to connect the communicable disease paradigm with the recently observed mortality patterns of the initial Coronavirus pandemic in March-April 2020 that have been shown by many to exhibit Gompertzian growth. In doing this, we show that this type of growth pattern is incompatible with traditional communicable disease spreading models, but rather with a model through which a ubiquitus and instanteneous stressor elicits a mutual stress response.
The findings of this paper call for a fundamental and interdisciplinary discussion of our accepted knowledge on communicable diseases, as the observations constitute a classic Kuhnian "anomaly" suggesting a paradigm shift away from that of a purely communicable disease paradigm to a hybrid where the environment plays a bigger role.}

@@

\subsection*{Unification through a modified SIR model}
It is also possible to reconcile the SIR model with the Gompertz curve. 
Inspired by the observation that in the linearized approximation there are only two coupled states, infected and susceptible, we augment the interaction term to higher orders. 
Then, we reverse the causality where the population of infected are now dependent on the population of the susceptible instead of the other way around as in (\ref{eq:networkSIR}),
\begin{equation}
\label{eq:reverseNetworkSIR}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(1-x_j) \quad \forall i.
\end{equation} 
This crucial change is based on the environmental stressor hypothesis rather than a communicable disease assumption.
Furthermore, we will for the sake of simplicity assume all nodes in the network have exactly one neighbor and that there exists a unique path between all nodes,
\begin{equation}
\sum_i a_{ij} = 1 \quad \forall j.
\end{equation} 
If we let $s_i = 1 - x_i$ be the probability of being susceptible, the augmented and causality-reversed SIR model infection term becomes 
\begin{equation}
\frac{d x_i}{dt} = \beta x_i\sum_j{a_{ij}}(s_{j} + s^2_j/2 + ...) - \alpha x_i\quad \forall i,
\end{equation}

@@

\begin{equation}
\frac{d x_i}{dt} = -\beta x_i\sum_j{a_{ij}}\ln{(1 - s_j)} - \alpha = -\beta x_i\sum_j{a_{ij}}\ln{x_j} - \alpha x_i \quad \forall i,
\end{equation}
As we are interested in the aggregate macroscopic behavior, we take averages in the log domain, exploit our setup where $\mathbf{A}$ is a single mapping from one node to another, and simplify to
\begin{equation}
\frac{d}{{dt}} \ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} = -\beta\ln\left[\prod_i^N{x_i}\right]^{\frac{1}{N}} - \frac{\alpha}{N}\sum_i^N x_i.
\end{equation}
If $\alpha\rightarrow 0$, then this equation will exhibit Gompertzian growth in the geometric ensemble average of microscopic units. 
An interpretation of $\alpha\rightarrow 0$ could be that the limiting factor emerges purely from the growth rate without the need for a second growth-limiting parameter, in line with the previous crucial hypothesis of causality reversal stated above. One further simplification could be seen in equating the logarithm of the geometric mean with the logarithm of the median of the set of $x_i$ to obtain (\ref{eq:medianGomp}),
\begin{equation}
\frac{d}{dt}\ln{M(t)} = -\beta \ln{M(t)}.
\end{equation}

@@

\matmethods{
\subsection*{Estimating model parameters}
The time-evolution of the SIRD and the Gompertz model are obtained with a Runge-Kutta algorithm, after first fitting parameters using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text (See SI appendix for details). 
Fitting is done using Python-Scipy's non-linear least squares algorithm with specified Jacobians and a 6-day windowed average of observed deaths, similar to the technique by Carletti et al. \citep{carletti2020covid}. 
Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.

% On statistical grounds, one could argue that the daily mortality counts follow a Poisson distribution. However, in a growth setting, one might expect heteroskedasticity inherent in the biological process where e.g. growth has less variance in the later stages, perhaps dependent on proximity to a stationary state. This latter hypothesis could be incorporated in the stochastic framework presented in the text, but to keep the scope of this paper confined we use a simple non-linear least squares minimization as our results are not meaningfully different compared with a more complex variance model.

### 2022-05-26 letterToEditor/letter.tex
update letter w refereese + slight mod to manuscript

\enclosure[Enclosed ]{
              \\ List of referees
%                  \hspace{0.0em} 5. GATE (2016) \hspace{2em} 9. Identity-Proof (Voter-Id)\\
%                  2. Diploma (MarkSheet/Certificate)
%                  \hspace{1.8em} 6. SLET (2021) \hspace{1.9em} 10. Resume\\        
%                  3. B.Tech (MarkSheet/Certificate)
%                  \hspace{2.3em} 7. UGC-NET (2022) \hspace{-0.1em} \\
%                  4. M.Tech (MarkSheet/Certificate)
%                  \hspace{2.1em} 8. Caste Certificate \\
                 }
\makelettertitle

@@

\makeletterclosing
\newpage
\textbf{List of possible referees}
\begin{enumerate}
\item N. El Karoui, mathematics, stochastic models
\item M. Molski, physics, Gompertz and quantum systems
\item W. Whitt, mathematics, stochastic models
\item N. C. Petroni, E. De Lauro, physics, Gompertz
\item E. B. Postnikov, physics, SIR models
\item G. Vattay, physics, SIR models
\item M. Levitt, computational biology, Gompertz models
\item E. Estrada, mathematics, graph theory, SIR models
\item T. Carletti, D. Fanelli, and F. Piazza, physics, SIR models
\item Z. Bajzer, biology, has a very good review paper on Gompertz modeling 
\item J. C. Mombach, physics, growth models
\item B. Shklovskii, physics, theory of Gompertz
\item D. Smilkov, C. or L. Kocarev, physics and mathematics, SIR models
\item C. G. Antonopoulos, physics, chaos and SIR models
\item K. Rypdal and M. Rypdal, mathematics, SIR models

>>>>

\enclosure[Enclosed ]{
              \\ List of referees, NAS members and PNAS editorial board members
%                  \hspace{0.0em} 5. GATE (2016) \hspace{2em} 9. Identity-Proof (Voter-Id)\\
%                  2. Diploma (MarkSheet/Certificate)
%                  \hspace{1.8em} 6. SLET (2021) \hspace{1.9em} 10. Resume\\        
%                  3. B.Tech (MarkSheet/Certificate)
%                  \hspace{2.3em} 7. UGC-NET (2022) \hspace{-0.1em} \\
%                  4. M.Tech (MarkSheet/Certificate)
%                  \hspace{2.1em} 8. Caste Certificate \\
                 }
\makelettertitle

@@

\makeletterclosing
\newpage

\textbf{List of Editorial board members}
\begin{enumerate}
\item Levine, Herbert
\item Levin, Simon A.
\item Marquet, Pablo A.
\end{enumerate}

\textbf{List of NAS members (in order of relevance)} 
\begin{enumerate}
\item Levine, Herbert
\item Levin, Simon A.
\item Michael Levitt

\item Nigel D. Goldenfeld 
\item Evans, Steven N.
\item Altshuler, Boris L.
\end{enumerate}
\textbf{List of possible referees}
\begin{enumerate}

\item Nicola Cufaro Petroni, cufaro@ba.infn.it,
Dipartimento di Matematica and TIRES, University of Bari (Ret).
Has done extensive work on the Gompertz model in relation to the Richard's model, connecting it with stochastic growth models and the Logistic (Lotka-Volterra) model. We use their concept of growth rooted in stochastic phenomena.
\item Salvatore De Martino,
demartino247@gmail.com, 
Dipartimento di Ingegneria dell’Informazione ed Elettrica.
Same as above
\item Marcin Molski, mamolski@amu.edu.pl, Adam Mickiewicz University, physics, Gompertz and quantum systems. Has done work on connecting the gompertz model to quantum coherent systems.
\item Francesco Piazza, Francesco.Piazza@cnrs-orleans.fr, Max Planck Institute for the Physics of Complex Systems, physics, SIR models.
Has written one of the central papers modeling covid disease propagation from the SIR perspective. We have based our methods on their paper.
\item Ernesto Estrada, estrada@ifisc.uib-csic.es, Institute of Interdisciplinary Physics and Complex Systems , mathematics, graph theory, SIR models.
Has written a recent paper on the Gompertz model in relation to communicable disease modelling including close connections with graph theoretical concepts used in our paper. 
\item K. Rypdal, kristoffer.rypdal@uit.no, University of Tromsø, mathematics, SIR models.
Has written a paper showing that SIR models become Gompertz in the extreme of the Richard's parameter. We also follow this paper's arguments in our paper.
\item Michael Levitt (NAS Member), michael.levitt@stanford.edu, Stanford University, computational biology, Gompertz models.
Has a preprint on the gompertz model in relation to covid where they leave the explanation of why mortality follows gompertz as an open question.
\item Chris G. Antonopoulos, canton@essex.ac.uk , U. of Essex, physics, chaos and SIR models.
Has written on the SIR model and chaotic dynamics.
\item Gabor Vattay, vattay@elte.hu, Eötvös Loránd University Budapest, physics, SIR models
Has written a paper on SIR models showing that there are ``initial transients'' in the observations which are explained by us in the Gompertz model.
\item E. B. Postnikov, postnikov@kursksu.ru, Kursk State Univesity, physics, SIR models.
Has done work on SIR models using a mean field approach, but has a broach research interest that could be a benefit when reviewing our paper.
\item Both H. Levine and Levin A. in the NAS list would be quite appropriate.
\item L. Kocarev, lkocarev@manu.edu.mk, Macedonian Academy of Sciences and Arts
, physics and mathematics, SIR models.
Has written papers on SIR models as they relate to non-markovian models.
